Automating Free Logic in HOL, with an Application in Category Theory

### 2017-04-26 FreeLogic/2017-Journal/article.tex
update

@@

\title{Automating Free Logic in HOL, with an Application in Category Theory\thanks{Mention DFG grants ...}}
%\title{Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\title{Embedding Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\subtitle{(And an Exemplary Application in Category Theory)}

@@

\author{Christoph Benzm\"uller       \and
      Dana Scott 
}

@@

\institute{F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           S. Author \at
              second address
}

@@

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers (and their integrated subprovers) for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is exemplarily employed in a selected domain of
  mathematics: starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issue in a prominent category
  theory textbook has been revealed.

@@

To demonstrate the practical relevance of our new system, we present,
in the second part of this article, a stepwise development of axiom
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be implemented and validated within modern proof assistants.

A total of eight different axiom systems is studied. The systems I-VI
are shown to be equivalent. The axiom system VII slightly modifies
axiom system VI to obtain (modulo notational transformation) the set
of axioms as proposed by Freyd and Scedrov in their textbook
``Categories, Allegories'' \cite{FreydScedrov90}, published in 1990;
see also Subsection \ref{subsec:FreydNotation} where we present their
original system.  While the axiom systems I-VI are shown to be
consistent, a constricted inconsistency result is obtained for system
VII (when encoded in free logic where free variables range over all
objects): We can prove $\exists x. \neg(\ex x) \Impl False$, where
$\ex$ is the existence predicate. Read this as: If there are undefined
objects, e.g. the value of an undefined composition $x\cdot y$
then we have falsity.  By contraposition, all objects (and thus all
compositions) must exist. But when we assume the latter, then the
axiom system VII essentially reduces categories to monoids.  We note
that axiom system V, which avoids this problem, corresponds to a set
of axioms proposed by Scott \cite{Scott79} in the 1970s. The
problem can also be avoided by restricting the variables in axiom
system VII to range only over existing objects and by postulating
strictness conditions.  This gives us axiom system VIII.

@@

We now formally introduce the syntax and semantics of free logic as
assumed in the remainder of this article. We refer to this logic with $\FFOL$.

@@

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x. r$ are (compound) formulas of $\FFOL$
\item given a formula $r$ of $\FFOL$, then $\Descr x. r$ is a
  term of $\FFOL$ (definite  description)
\end{enumerate}
\end{definition}

@@

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a triple
  $\langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{I,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{I,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{I,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{I,g} = I(f)(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x. r \|^{I,g} = d \in \ex$, such that $\| r
  \|^{I,g[d/x]} = \True$ and  $\| r
  \|^{I,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x. r \|^{I,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| q \|^{I,g} = I(q)$, where $q\in\P$ is an $0$-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{I,g} = \True$ iff $\| t_1 \|^{I,g} = \| t_2
  \|^{I,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in Section \ref{??}, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{I,g} = \True$ iff $(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
\item $\| r \Impl s \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
  or $\| s \|^{I,g} = \True$
\item $\| \forall x. r  \|^{I,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{I,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\begin{definition}\label{homlvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ iff
  $\|s_o\|^{M,g} = T$; this is also denoted as $M,g \models s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models s_o$, iff $M,g \models s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models s_o$, iff $s_o$ is valid for all $M$. Moreover,
  we write $\Gamma\models \Delta$ (for sets of formulas $\Gamma$ and
  $\Delta$) iff there is a model $M$ and an assignment $g$ such that
  $M,g \models s_o$ for all $s_o\in \Gamma$ and $M,g \models t_o$ for
  at least one $t_o\in \Delta$.
\end{definition}

@@

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ is defined by the following
grammar:\footnote{Ednote: Maybe better to define this analagous to $\FFOL$.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\, t_\alpha)_\beta} \mid  \hol{\neg_{o\ar o}\, s_o} \mid \\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$) and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha} s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in the usual way.
\end{definition}

@@

\begin{definition}\label{homlframe}
A \emph{frame} $\hol{D}$ is a collection $\hol{\{D_\alpha\}_{\alpha\in\entity{T}}}$
of nonempty sets $\hol{D_\alpha}$, such that $\hol{D_o = \{T,F\}}$
(for truth and falsehood).  The
$\hol{D_{\alpha\ar\beta}}$ are collections of functions mapping
$\hol{D_\alpha}$ into $\hol{D_\beta}$. 
\end{definition}

\begin{definition}\label{holmodel}
  A \emph{model} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{definition}

\begin{definition}\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = \hol{T}$ iff $\hol{\|s_o\|^{M,g}} = \hol{F}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  \hol{T}$ iff\, $\hol{\|s_o\|^{M,g}} = \hol{T}$ or $\hol{\|t_o\|^{M,g}}
  = \hol{T}$
\item $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{T}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

\begin{definition}\label{homlhenkinmodel}
 A model $\hol{M}=\hol{\langle D, I \rangle}$ is called a
 \emph{standard model} iff for all $\hol{\alpha,\beta}\in \hol{T}$ we have
 $\hol{D_{\alpha\ar\beta}} = \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$. In a \emph{Henkin model} function spaces are not necessarily
 full. Instead it is only required that $\hol{D_{\alpha\ar\beta}}
 \subseteq \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$ (for all
 $\hol{\alpha,\beta}\in \hol{T}$) and that the valuation function 
 $\hol{\|\cdot\|^{M,g}}$ from above is total (i.e., every term denotes). Any standard model is obviously
 also a Henkin model. We consider Henkin models in the remainder.
\end{definition}

\begin{definition}\label{homlvalid}
A formula $\hol{s_o}$ is \emph{true} in model $M$ 
under assignment $\hol{g}$ iff $\hol{\|s_o\|^{M,g}} = \hol{T}$; this is also denoted
as $\hol{M,g \models^\HOL s_o}$.  A formula $\hol{s_o}$ is called \emph{valid} in
$\hol{M}$, which is denoted as $\hol{M \models^\HOL s_o}$, iff $\hol{M,g \models^\HOL s_o}$ for all
assignments $\hol{g}$. Finally, a formula $\hol{s_o}$ is called
\emph{valid}, which we denote by $\hol{\models^\HOL s_o}$, iff $\hol{s_o}$ is valid for
all $\hol{M}$. Moreover, we write $\hol{\Gamma\models^\HOL \Delta}$ (for sets of
 formulas $\hol{\Gamma}$ and $\hol{\Delta}$) iff there is a model $\hol{M}$ and an assignment $\hol{g}$ such that  
 $\hol{M,g \models^\HOL s_o}$ for all $\hol{s_o}\in \hol{\Gamma}$ and $\hol{M,g \models^\HOL t_o}$ for at least one $\hol{t_o}\in \hol{\Delta}$.
\end{definition}

@@

>>>>

%\usepackage{authblk}

@@

\title{Automating Free Logic in HOL, with an Application in Category Theory}
%\title{Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\title{Embedding Free Logic and Axiom Systems for Category Theory in HOL\thanks{Mention DFG grants ...}}
%\subtitle{(And an Exemplary Application in Category Theory)}

@@

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   University of Luxembourg, Luxembourg \& Freie Universit\"at Berlin, Berlin, Germany \\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of Califormia, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

\date{\today}                     %% if you don't need date to appear

% \author{Christoph Benzm\"uller       \and
%       Dana Scott 
% }

@@

% \institute{F. Author \at
%               first address \\
%               Tel.: +123-45-678910\\
%               Fax: +123-45-678910\\
%               \email{fauthor@example.com}           %  \\
% %             \emph{Present address:} of F. Author  %  if needed
%            \and
%            S. Author \at
%               second address
% }

@@

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is exemplarily employed in a selected domain of
  mathematics: starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issue in a prominent category
  theory textbook has been revealed.

@@

To demonstrate the practical relevance of our new system, we present
in \S4 a stepwise development of axioms
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be implemented and validated within modern proof assistants such as
Isabelle/HOL.  We also address the relation of our axioms systems to
alternative proposals from the literature, including an axiom set
proposed by Freyd and Scedrov in their textbook ``Categories,
Allegories'' \cite{FreydScedrov90} for which we reveal a technical
flaw: either all operations, e.g. morphism composition, are total or
their axiom system is inconsistent. The repair for this problem is
quite straightforward, however. The solution essentially corresponds
to a set of axioms proposed by Scott \cite{Scott79} in the 1970s.

% \marginpar{will rewrite this part}
% A total of eight different axiom systems is studied. The systems I-VI
% are shown to be equivalent. The axiom system VII slightly modifies
% axiom system VI to obtain (modulo notational transformation) the set
% of axioms as proposed by Freyd and Scedrov in their textbook
% ``Categories, Allegories'' \cite{FreydScedrov90}, published in 1990;
% see also Subsection \ref{subsec:FreydNotation} where we present their
% original system.  While the axiom systems I-VI are shown to be
% consistent, a constricted inconsistency result is obtained for system
% VII (when encoded in free logic where free variables range over all
% objects): We can prove $\exists x\lambdot \neg(\ex x) \Impl False$, where
% $\ex$ is the existence predicate. Read this as: If there are undefined
% objects, e.g. the value of an undefined composition $x\cdot y$
% then we have falsity.  By contraposition, all objects (and thus all
% compositions) must exist. But when we assume the latter, then the
% axiom system VII essentially reduces categories to monoids.  We note
% that axiom system V, which avoids this problem, corresponds to a set
% of axioms proposed by Scott \cite{Scott79} in the 1970s. The
% problem can also be avoided by restricting the variables in axiom
% system VII to range only over existing objects and by postulating
% strictness conditions.  This gives us axiom system VIII.

@@

We now formally introduce the syntax and semantics of Scott's \cite{Scott67} free logic as
assumed in the remainder of this article. We refer to this logic with $\FFOL$.

@@

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x\lambdot r$ are (compound) formulas of $\FFOL$
\item given a formula $r$ of $\FFOL$, then $\Descr x\lambdot r$ is a
  term of $\FFOL$ (definite  description)
\end{enumerate}
\end{definition}

@@

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{M,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| q \|^{M,g} = I(q)$, where $q\in\P$ is an $0$-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{M,g} = \True$ iff $\| t_1 \|^{M,g} = \| t_2
  \|^{M,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in \S4, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{M,g} = \True$ iff $(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{M,g} = \True$ iff $\| r \|^{M,g} = \False$
\item $\| r \Impl s \|^{M,g} = \True$ iff $\| r \|^{M,g} = \False$
  or $\| s \|^{M,g} = \True$
\item $\| \forall x\lambdot r  \|^{M,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{M,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\begin{definition}[Validity]\label{ffollvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ iff
  $\|s_o\|^{M,g} = T$; this is also denoted as $M,g \models^\FFOL s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models^\FFOL s_o$, iff $M,g \models^\FFOL s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models^\FFOL s_o$, iff $s_o$ is valid for all $M$. Moreover,
  we write $\Gamma\models^\FFOL \Delta$ (for sets of formulas $\Gamma$ and
  $\Delta$) iff there is a model $M$ and an assignment $g$ such that
  $M,g \models^\FFOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\FFOL t_o$ for
  at least one $t_o\in \Delta$.
\end{definition}

@@

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ with primitive equality and
choice is defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \cite{Andrews02}. The choice here is motivated by the
  signature supported in higher-order automated theorem provers such
  as LEO-II \cite{LEO-II} and Leo-III \cite{Leo-III}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$) and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

@@

\begin{definition}[Frame for \HOL]\label{homlframe}
A \emph{frame} $\hol{D}$ is a collection $\hol{\{D_\alpha\}_{\alpha\in\entity{T}}}$
of nonempty sets $\hol{D_\alpha}$, such that $\hol{D_o} = \{\True,\False\}$
(where, as before, $\True$ and $\False$ denote truth and falsehood).  The
$\hol{D_{\alpha\ar\beta}}$ are collections of functions mapping
$\hol{D_\alpha}$ into $\hol{D_\beta}$. 
\end{definition}

\begin{definition}[Model for \HOL]\label{holmodel}
  A \emph{model} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{definition}

\begin{definition}[Evaluation Function]\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|s_\alpha = t_\alpha\|^{M,g}} = {\True}$ iff $\hol{\|s_\alpha\|^{M,g}} = \hol{\|t_\alpha\|^{M,g}} $
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = {\True}$ iff $\hol{\|s_o\|^{M,g}} = {\False}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  {\True}$ iff\, $\hol{\|s_o\|^{M,g}} = {\True}$ or $\hol{\|t_o\|^{M,g}}
  = {\True}$
\item \sloppy $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = {\True}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\False}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

\begin{definition}[Standard model and Henkin model]\label{homlhenkinmodel}
 A model $\hol{M}=\hol{\langle D, I \rangle}$ is called a
 \emph{standard model} iff for all $\hol{\alpha,\beta}\in \hol{T}$ we have
 $\hol{D_{\alpha\ar\beta}} = \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$. In a \emph{Henkin model} function spaces are not necessarily
 full. Instead it is only required that $\hol{D_{\alpha\ar\beta}}
 \subseteq \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$ (for all
 $\hol{\alpha,\beta}\in \hol{T}$) and that the valuation function 
 $\hol{\|\cdot\|^{M,g}}$ from above is total (i.e., every term denotes). Any standard model is obviously
 also a Henkin model. We consider Henkin models in the remainder.
\end{definition}

\begin{definition}[Validity]\label{homlvalid}
A formula $\hol{s_o}$ is \emph{true} in model $M$ 
under assignment $\hol{g}$ iff $\hol{\|s_o\|^{M,g}} = {\True}$; this is also denoted
as $\hol{M,g \models^\HOL s_o}$.  A formula $\hol{s_o}$ is called \emph{valid} in
$\hol{M}$, which is denoted as $\hol{M \models^\HOL s_o}$, iff $\hol{M,g \models^\HOL s_o}$ for all
assignments $\hol{g}$. Finally, a formula $\hol{s_o}$ is called
\emph{valid}, which we denote by $\hol{\models^\HOL s_o}$, iff $\hol{s_o}$ is valid for
all $\hol{M}$. Moreover, we write $\hol{\Gamma\models^\HOL \Delta}$ (for sets of
 formulas $\hol{\Gamma}$ and $\hol{\Delta}$) iff there is a model $\hol{M}$ and an assignment $\hol{g}$ such that  
 $\hol{M,g \models^\HOL s_o}$ for all $\hol{s_o}\in \hol{\Gamma}$ and $\hol{M,g \models^\HOL t_o}$ for at least one $\hol{t_o}\in \hol{\Delta}$.
\end{definition}

@@

%\pagebreak

In the remainder of this article we present an application of our
framework for the exploration of axioms systems in category theory. In
this context description is not required and we hence omit it in our 
theory. Figure \ref{pic1} shows the embedding of $\FFOL$ in $\HOL$,
encoded in Isabelle/HOL, 
that we have employed in our experiments. A related Isabelle/HOL
encoding of the above theory with description has been presented and  discussed in
\cite{ICMS}.
